Comparative Analysis of Lion and AdamW Optimizers for Cross-Encoder Reranking with MiniLM, GTE, and ModernBERT

### 2025-06-13 cross_encoder/thesis/Chapters/2.literature.tex
Implement code changes to enhance functionality and improve performance

\section{Research Gaps}

While cross-encoder effectiveness is well-established, systematic investigation of optimization strategies remains limited. Most studies assume AdamW without exploring alternatives like Lion, particularly for information retrieval tasks where ranking precision is critical.

\subsection{Modern Transformer Architectures}

Recent developments in transformer architectures have introduced models specifically designed for improved efficiency and longer context processing. The General Text Embeddings (GTE) family \cite{li2023towards} demonstrated that models trained on diverse, multilingual corpora could achieve strong performance across various text understanding tasks, including retrieval and reranking.

ModernBERT \cite{modernbert} represents the current state-of-the-art in encoder-only transformers, incorporating several architectural innovations:

\begin{itemize}
    \item \textbf{Rotary Positional Embeddings (RoPE)} \cite{su2023roformerenhancedtransformerrotary}: Enable effective handling of longer sequences by providing more flexible positional encoding.
    \item \textbf{Flash Attention} \cite{dao2022flashattentionfastmemoryefficientexact}: Improves memory efficiency and computational speed for attention operations.
    \item \textbf{GeGLU Activation Functions} \cite{shazeer2020gluvariantsimprovetransformer}: Enhance model expressiveness while maintaining computational efficiency.
\end{itemize}

These architectural improvements enable ModernBERT to process sequences up to 8192 tokens efficiently, making it particularly suitable for long document processing tasks.

%----------------------------------------------------------------------------------------
\section{Optimization Algorithms for Deep Learning}

The success of neural information retrieval models depends heavily on the optimization algorithms used during training. The choice of optimizer affects convergence speed, final performance, and training stability. This section reviews the evolution of optimization algorithms and their specific applications to transformer-based models.

\subsection{Classical Optimization Methods}

Stochastic Gradient Descent (SGD) remains a fundamental optimization algorithm, providing theoretical guarantees and interpretable behavior. However, the challenges of training deep neural networks - including vanishing gradients, saddle points, and varying curvature across parameter space - motivated the development of adaptive optimization methods.

The introduction of momentum to SGD addressed some limitations by accumulating gradients across iterations, helping optimization navigate through areas of high curvature and accelerate convergence in consistent directions. However, SGD with momentum still struggled with the varying scales of different parameters and the need for careful learning rate tuning.

\subsection{Adaptive Learning Rate Methods}

AdaGrad \cite{duchi2011adaptive} introduced the concept of adaptive learning rates, automatically adjusting step sizes based on historical gradient information. By maintaining per-parameter learning rates inversely proportional to the square root of accumulated squared gradients, AdaGrad could handle sparse features effectively and reduce the need for manual learning rate tuning.

RMSprop \cite{tieleman2012lecture} addressed AdaGrad's aggressive learning rate decay by using exponential moving averages instead of accumulating all historical gradients. This modification prevented the learning rate from becoming too small too quickly, enabling continued learning throughout training.

Adam \cite{kingma2017adam} combined the benefits of momentum-based methods with adaptive learning rates, maintaining separate exponentially decaying averages for both gradients (first moment) and squared gradients (second moment). The Adam optimizer became widely adopted due to its robustness across different architectures and tasks, requiring minimal hyperparameter tuning in many scenarios.

\subsection{AdamW and Weight Decay Regularization}

AdamW \cite{loshchilov2019decoupled} represented a significant improvement over Adam by decoupling weight decay regularization from the adaptive learning rate mechanism. The key insight was that traditional L2 regularization in Adam led to suboptimal behavior because the adaptive learning rate scaling affected both gradient and regularization terms.

By separating weight decay from gradient-based updates, AdamW achieved better generalization performance and more stable training dynamics. The decoupled formulation:

\begin{equation}
\theta_{t+1} = \theta_t - \eta \left( \frac{\hat{m}_t}{\sqrt{\hat{v}_t} + \epsilon} + \lambda \theta_t \right)
\end{equation}

where $\lambda$ represents the weight decay coefficient applied directly to parameters, independent of the adaptive learning rate scaling.

AdamW quickly became the standard optimizer for transformer-based models, demonstrating superior performance across various natural language processing tasks and establishing itself as the default choice for most deep learning applications.

\subsection{Lion Optimizer: A Novel Approach}

The Lion (EvoLved Sign mOmeNtum) optimizer \cite{chen2023symbolic} represents a departure from traditional adaptive optimization approaches. Developed through symbolic mathematics and program search, Lion utilizes a simplified update rule that relies primarily on the sign of momentum-based gradients.

The Lion update mechanism follows:

\begin{align}
c_t &= \beta_1 m_{t-1} + (1 - \beta_1) g_t \\
\theta_t &= \theta_{t-1} - \eta \left( \text{sign}(c_t) + \lambda \theta_{t-1} \right) \\
m_t &= \beta_2 m_{t-1} + (1 - \beta_2) g_t
\end{align}

>>>>

\section{Summary}

This literature review establishes the foundation for understanding cross-encoder architectures, optimization algorithms, and evaluation methodologies in information retrieval. The reviewed work demonstrates the effectiveness of transformer-based cross-encoders and the dominance of AdamW as the standard optimizer. However, the systematic exploration of alternative optimization strategies, particularly the recently proposed Lion optimizer, remains an open area for investigation in the context of information retrieval tasks.

%----------------------------------------------------------------------------------------

### 2025-06-13 cross_encoder/thesis/Chapters/3.research_gap_and_proposed_solution.tex
Implement code changes to enhance functionality and improve performance

While AdamW has become standard for transformer training, the Lion optimizer's effectiveness for cross-encoder IR tasks remains unexplored. Key gaps include:

\begin{itemize}
    \item \textbf{Novel Optimizer Evaluation}: Lion optimizer's memory efficiency and simplified updates haven't been systematically evaluated for cross-encoder training in IR contexts.
    \item \textbf{Architecture-Optimizer Interaction}: Different transformer architectures (MiniLM, GTE, ModernBERT) may respond differently to optimization strategies.
    \item \textbf{Training Efficiency Trade-offs}: Lion's claimed efficiency advantages need validation for computationally intensive cross-encoder training.
    \item \textbf{Comprehensive IR Benchmarking}: Systematic evaluation across standard IR datasets (MS MARCO, TREC DL) with multiple metrics is needed.
\end{itemize}

@@

This research systematically compares Lion and AdamW optimizers across three cross-encoder architectures for information retrieval tasks.

\subsection{Model Selection}
Three transformer architectures representing different efficiency-effectiveness points:
\begin{itemize}
    \item \textbf{MiniLM-L12-H384}: Compact model (384 hidden dims, 12 layers) for efficiency
    \item \textbf{GTE-multilingual-base}: Multilingual model with 8192 token context support
    \item \textbf{ModernBERT-base}: State-of-the-art encoder with RoPE, Flash Attention, GeGLU
\end{itemize}

\subsection{Experimental Design}
\textbf{Training Setup}: Modal cloud platform with NVIDIA A100-80GB GPUs, batch size 16, 3 epochs, MS MARCO passage dataset.

\textbf{Optimizer Configurations}:
\begin{itemize}
    \item \textbf{AdamW}: Learning rates 2e-5 (MiniLM, GTE), 2e-6 (ModernBERT), weight decay 0.01
    \item \textbf{Lion}: Same learning rates, β₁=0.9, β₂=0.99, sign-based momentum updates
\end{itemize}

\textbf{Evaluation}: TREC 2019 DL Track and MS MARCO dev set using NDCG@10, MAP, MRR@10 metrics.

>>>>

Despite the established effectiveness of cross-encoder models for information retrieval, several critical gaps exist in current research that motivate this investigation:

\subsection{Limited Optimization Strategy Exploration}

While AdamW has become the de facto standard for transformer training, the systematic investigation of optimization strategies for cross-encoder models remains limited. Most studies assume AdamW without exploring alternatives, particularly for information retrieval tasks where ranking precision is critical. The recently proposed Lion optimizer \cite{chen2023symbolic}, with its novel sign-based momentum approach, presents an unexplored opportunity for cross-encoder optimization.

\subsection{Architecture-Optimizer Interaction Analysis}

Different transformer architectures may respond differently to optimization strategies, yet this interaction has not been systematically studied. Modern architectures like:

\begin{itemize}
    \item \textbf{ModernBERT} \cite{modernbert}: Incorporating Rotary Positional Embeddings (RoPE), Flash Attention, and GeGLU activations with 8192 token context support
    \item \textbf{GTE models} \cite{li2023towards}: Multilingual architectures trained on diverse corpora with extended context capabilities
    \item \textbf{MiniLM} \cite{wang2020minilm}: Efficient distilled variants designed for computational efficiency
\end{itemize}

Each may exhibit different optimization dynamics with various optimizers, requiring systematic evaluation.

\subsection{Information Retrieval Context Specificity}

Most optimizer comparisons focus on language modeling or general NLP tasks. Information retrieval presents unique challenges:
\begin{itemize}
    \item \textbf{Ranking Quality Focus}: IR metrics prioritize ranking precision over classification accuracy
    \item \textbf{Query-Document Interaction}: Cross-encoder architectures require different optimization considerations than standard sequence models
    \item \textbf{Training Efficiency}: Large-scale IR datasets demand memory-efficient training approaches
\end{itemize}

\subsection{Modern Transformer Architecture Gaps}

Recent architectural innovations have not been thoroughly evaluated with alternative optimizers:

\subsubsection{ModernBERT Architectural Advances}
ModernBERT \cite{modernbert} represents the current state-of-the-art in encoder-only transformers, incorporating several key innovations:

\begin{itemize}
    \item \textbf{Rotary Positional Embeddings (RoPE)} \cite{su2023roformerenhancedtransformerrotary}: Enable effective handling of longer sequences up to 8192 tokens by providing more flexible positional encoding compared to traditional absolute positional embeddings.
    \item \textbf{Flash Attention} \cite{dao2022flashattentionfastmemoryefficientexact}: Improves memory efficiency and computational speed for attention operations, particularly beneficial for longer sequences.
    \item \textbf{GeGLU Activation Functions} \cite{shazeer2020gluvariantsimprovetransformer}: Enhance model expressiveness while maintaining computational efficiency compared to traditional ReLU activations.
\end{itemize}

These architectural improvements enable efficient processing of long documents but their interaction with different optimizers, particularly Lion's memory-efficient approach, remains unexplored.

\subsubsection{GTE Model Family}
The General Text Embeddings (GTE) family \cite{li2023towards} demonstrated that models trained on diverse, multilingual corpora could achieve strong performance across various text understanding tasks. The GTE-multilingual-base model supports 75 languages and 8192 token context length, but optimization strategy comparisons for this architecture in IR contexts are lacking.

\subsection{Comprehensive IR Benchmarking Gap}

Existing optimizer evaluations typically focus on limited metrics or datasets. A comprehensive evaluation across standard IR benchmarks (MS MARCO, TREC DL) with multiple effectiveness metrics (NDCG@10, MAP, MRR@10, Recall@10, R-Prec, P@10) is needed to establish optimizer performance in IR contexts.

\subsection{Training Efficiency and Memory Usage Analysis}

Lion optimizer claims memory efficiency advantages due to its simplified momentum tracking, but these benefits have not been validated for computationally intensive cross-encoder training scenarios typical in information retrieval applications.

@@

This research addresses the identified gaps through a systematic comparative study of Lion and AdamW optimizers across three distinct cross-encoder architectures for information retrieval tasks.

\subsection{Research Objectives}

\begin{itemize}
    \item \textbf{Primary Objective}: Compare Lion and AdamW optimizer effectiveness for cross-encoder reranking across multiple transformer architectures
    \item \textbf{Secondary Objectives}: 
    \begin{itemize}
        \item Analyze architecture-optimizer interaction patterns
        \item Evaluate training dynamics and convergence behavior
        \item Assess memory efficiency and computational requirements
        \item Provide comprehensive IR benchmark results
    \end{itemize}
\end{itemize}

\subsection{Model Selection Rationale}

Three transformer architectures representing different efficiency-effectiveness trade-offs:

\subsubsection{MiniLM-L12-H384-uncased}
\begin{itemize}
    \item \textbf{Architecture}: Compact model (384 hidden dimensions, 12 layers)
    \item \textbf{Design Focus}: Efficiency through knowledge distillation from larger BERT models
    \item \textbf{Context Length}: Standard 512 tokens
    \item \textbf{Use Case}: Resource-constrained environments requiring fast inference
\end{itemize}

\subsubsection{GTE-multilingual-base}
\begin{itemize}
    \item \textbf{Architecture}: Multilingual encoder with transformer++ backbone (BERT + RoPE + GLU)
    \item \textbf{Context Support}: 8192 tokens for long document processing
    \item \textbf{Language Coverage}: 75+ languages with strong cross-lingual capabilities
    \item \textbf{Training Data}: Diverse multilingual corpora for robust text understanding
\end{itemize}

\subsubsection{ModernBERT-base}
\begin{itemize}
    \item \textbf{Architecture}: State-of-the-art encoder with advanced optimizations
    \item \textbf{Key Features}: RoPE, Flash Attention, GeGLU activations
    \item \textbf{Context Length}: 8192 tokens with efficient long-context processing
    \item \textbf{Training Scale}: 2 trillion tokens for comprehensive language understanding
\end{itemize}

\subsection{Experimental Design}

\subsubsection{Training Configuration}

\textbf{Infrastructure}: Modal cloud platform with NVIDIA A100-80GB GPUs providing consistent computational environment

\textbf{Training Parameters}:
\begin{itemize}
    \item \textbf{Batch Size}: 16 per GPU for memory efficiency
    \item \textbf{Training Duration}: 3 epochs with checkpoint evaluation after each epoch
    \item \textbf{Dataset}: MS MARCO passage ranking triplets (~2M query-passage pairs)
    \item \textbf{Precision}: BF16 for improved training efficiency
    \item \textbf{Random Seed}: 12 for reproducible results
\end{itemize}

\subsubsection{Optimizer Configuration Details}

\textbf{AdamW Configuration}:
\begin{itemize}
    \item \textbf{Learning Rates}: 
    \begin{itemize}
        \item 2e-5 for MiniLM and GTE (standard rate)
        \item 2e-5 for ModernBERT (baseline comparison)
    \end{itemize}
    \item \textbf{Weight Decay}: 0.01 for regularization
    \item \textbf{Beta Parameters}: β₁=0.9, β₂=0.999 (default values)
    \item \textbf{Scheduler}: Linear warmup for 10\% of training steps
\end{itemize}

\textbf{Lion Configuration}:
\begin{itemize}
    \item \textbf{Learning Rates}: 
    \begin{itemize}
        \item 2e-5 for MiniLM and GTE (consistency with AdamW)
        \item 2e-6 for ModernBERT (optimized based on architecture characteristics)
    \end{itemize}
    \item \textbf{Beta Parameters}: β₁=0.9, β₂=0.99 (recommended values)
    \item \textbf{Weight Decay}: 0.01 for fair comparison
    \item \textbf{Scheduler}: Cosine Annealing for ModernBERT, linear for others
\end{itemize}

\subsection{Evaluation Protocol}

\subsubsection{Datasets and Metrics}

\textbf{Primary Evaluation}: TREC 2019 Deep Learning Track passage ranking
\begin{itemize}
    \item \textbf{Query Set}: 43 queries with graded relevance judgments
    \item \textbf{Candidate Retrieval}: Top-1000 BM25 candidates per query
    \item \textbf{Metrics}: NDCG@10, MAP, Recall@10, R-Precision, P@10
\end{itemize}

\textbf{Secondary Evaluation}: MS MARCO passage development set
\begin{itemize}
    \item \textbf{Focus Metric}: MRR@10 (Mean Reciprocal Rank at cutoff 10)
    \item \textbf{Purpose}: Complementary perspective on larger query set
\end{itemize}

\subsubsection{Evaluation Framework}

\begin{itemize}
    \item \textbf{Reranking Process}: Cross-encoder scoring of BM25 candidate sets
    \item \textbf{Evaluation Tools}: Standard trec\_eval for consistent metric calculation
    \item \textbf{Checkpoint Strategy}: Evaluation after each training epoch
    \item \textbf{Performance Tracking}: Weights \& Biases for experiment monitoring
\end{itemize}

\subsection{Controlled Variables and Experimental Validity}

\begin{itemize}
    \item \textbf{Hardware Consistency}: All experiments on identical A100-80GB GPUs
    \item \textbf{Software Environment}: Standardized Modal containers with fixed library versions
    \item \textbf{Data Consistency}: Identical training and evaluation data across all configurations
    \item \textbf{Random Seed Control}: Fixed seed for reproducible initialization and data shuffling
    \item \textbf{Hyperparameter Documentation}: Complete parameter logging for replication
\end{itemize}

### 2025-06-13 cross_encoder/thesis/Chapters/6.summary_and_future_scope_of_work.tex
Implement code changes to enhance functionality and improve performance

This chapter presents a comprehensive summary of our research findings on optimizer effectiveness in cross-encoder reranking and outlines promising directions for future investigation. We reflect on the key insights gained from our experimental analysis and discuss potential avenues for extending this work.

\section{Summary of Research}

Our investigation into the comparative effectiveness of Lion and AdamW optimizers for cross-encoder reranking has yielded several significant findings:

\subsection{Key Findings}

\begin{enumerate}
    \item \textbf{Optimizer-Model Interactions:}
    \begin{itemize}
        \item The effectiveness of optimizers showed strong dependence on model architecture and training configuration
        \item ModernBERT achieved optimal performance with Lion optimizer using specialized learning rate settings
        \item GTE demonstrated superior performance with AdamW under standard training parameters
        \item MiniLM showed varying preferences between optimizers depending on the evaluation metric
    \end{itemize}

    \item \textbf{Performance Achievements:}
    \begin{itemize}
        \item ModernBERT with Lion achieved best-in-class results:
        \begin{itemize}
            \item NDCG@10: 0.7225
            \item MAP: 0.5115
            \item R-Precision: 0.5183
        \end{itemize}
        \item Both MiniLM and ModernBERT with Lion achieved state-of-the-art MRR@10 (0.5988) on MS MARCO dev
    \end{itemize}

    \item \textbf{Training Dynamics:}
    \begin{itemize}
        \item Lion demonstrated more stable evaluation loss curves
        \item Cosine Annealing learning rate schedule proved particularly effective with Lion
        \item Gradient behavior showed distinct patterns between optimizers
    \end{itemize}
\end{enumerate}

\subsection{Technical Insights}

The research revealed several important technical considerations:

\begin{itemize}
    \item \textbf{Learning Rate Sensitivity:} The dramatic impact of learning rate selection on Lion's performance suggests careful tuning is essential
    \item \textbf{Context Length Benefits:} Models supporting longer contexts (8192 tokens) showed generally superior performance
    \item \textbf{Architecture Synergies:} Modern architectural features (RoPE, Flash Attention) appeared to complement Lion's optimization characteristics
\end{itemize}

@@

Our findings open several promising avenues for future research:

\subsection{Technical Extensions}

\begin{enumerate}
    \item \textbf{Hyperparameter Optimization:}
    \begin{itemize}
        \item Comprehensive grid search for Lion's optimal parameters across different model sizes
        \item Investigation of alternative learning rate schedules
        \item Exploration of momentum parameter impacts
    \end{itemize}

    \item \textbf{Architecture Studies:}
    \begin{itemize}
        \item Evaluation of Lion's effectiveness on emerging transformer variants
        \item Investigation of optimization patterns in multi-query attention mechanisms
        \item Analysis of position embedding schemes' interaction with different optimizers
    \end{itemize}

    \item \textbf{Scaling Studies:}
    \begin{itemize}
        \item Analysis of optimizer behavior with larger model sizes
        \item Investigation of training stability at different batch sizes
        \item Evaluation of memory efficiency at scale
    \end{itemize}
\end{enumerate}

\subsection{Application Extensions}

Several practical applications deserve further investigation:

\begin{itemize}
    \item \textbf{Document-Level Reranking:} Leveraging the 8K context capability for full document reranking
    \item \textbf{Multi-Stage Ranking:} Investigating optimizer impact in cascade ranking architectures
    \item \textbf{Cross-Lingual Applications:} Extending the analysis to multilingual reranking scenarios
    \item \textbf{Domain Adaptation:} Studying optimizer effectiveness in domain transfer settings
\end{itemize}

\subsection{Theoretical Investigations}

Future work should also address theoretical aspects:

\begin{itemize}
    \item Analysis of Lion's convergence properties in ranking optimization
    \item Mathematical characterization of optimizer-architecture interactions
    \item Theoretical bounds on performance with different optimizers
\end{itemize}

\section{Recommendations for Practitioners}

Based on our findings, we recommend:

\begin{itemize}
    \item Consider Lion optimizer for modern architectures with appropriate learning rate scheduling
    \item Maintain AdamW as a robust baseline, especially for established architectures
    \item Carefully tune learning rates when using Lion optimizer
    \item Monitor training dynamics through multiple metrics for optimal checkpoint selection
\end{itemize}

\section{Concluding Remarks}

This research has demonstrated the potential of the Lion optimizer in cross-encoder reranking while highlighting the complexity of optimizer-model interactions. The findings provide a foundation for both practical applications and future research directions in neural information retrieval.

@@

We gratefully acknowledge Modal Labs (\url{https://modal.com/}) for providing the cloud computing infrastructure and GPU resources (NVIDIA A100-80GB) that made this research possible. Their platform enabled efficient experimentation with large-scale models and datasets, contributing significantly to the comprehensive nature of our analysis.

>>>>

This chapter summarizes the key contributions of our research and outlines promising directions for future investigation in optimizer effectiveness for cross-encoder reranking.

\section{Research Summary}

This thesis investigated the comparative effectiveness of Lion and AdamW optimizers for cross-encoder reranking across three transformer architectures: MiniLM, GTE, and ModernBERT.

\subsection{Primary Contributions}

\begin{enumerate}
    \item \textbf{First Systematic Evaluation}: Conducted the first comprehensive comparison of Lion optimizer for cross-encoder information retrieval tasks across multiple architectures.

    \item \textbf{Architecture-Optimizer Interaction Analysis}: Demonstrated that optimizer effectiveness varies significantly based on model architecture and training configuration.

    \item \textbf{State-of-the-art Results}: Achieved best-in-class performance with ModernBERT + Lion (NDCG@10: 0.7225) and tied best MRR@10 (0.5988) on standard IR benchmarks.

    \item \textbf{Training Dynamics Insights}: Provided detailed analysis of convergence patterns and gradient behavior differences between optimizers.
\end{enumerate}

\subsection{Key Findings Summary}

Our experimental results establish that:
\begin{itemize}
    \item \textbf{No Universal Optimizer}: Neither Lion nor AdamW consistently outperforms across all architectures
    \item \textbf{Configuration Sensitivity}: Optimizer effectiveness depends heavily on learning rate and scheduling choices
    \item \textbf{Modern Architecture Benefits}: Advanced transformer features synergize well with Lion's optimization approach
    \item \textbf{Metric-Dependent Performance}: Different optimizers excel on different evaluation metrics
\end{itemize}

@@

This research opens several promising avenues for future investigation:

\subsection{Immediate Extensions}

\begin{enumerate}
    \item \textbf{Comprehensive Hyperparameter Studies}:
    \begin{itemize}
        \item Systematic grid search for Lion's optimal parameters across model architectures
        \item Investigation of alternative learning rate schedules beyond Cosine Annealing
        \item Analysis of momentum parameter sensitivity
    \end{itemize}

    \item \textbf{Extended Architecture Coverage}:
    \begin{itemize}
        \item Evaluation on larger model sizes (large and extra-large variants)
        \item Testing on emerging transformer architectures (e.g., Mamba, RetNet)
        \item Investigation of decoder-only models for reranking tasks
    \end{itemize}

    \item \textbf{Enhanced Evaluation Scope}:
    \begin{itemize}
        \item Extended evaluation on additional IR benchmarks (BEIR, LoTTE)
        \item Cross-domain evaluation for generalization assessment
        \item Multi-language reranking performance analysis
    \end{itemize}
\end{enumerate}

\subsection{Advanced Research Directions}

\begin{itemize}
    \item \textbf{Theoretical Analysis}: Mathematical characterization of Lion's convergence properties in ranking optimization contexts

    \item \textbf{Memory and Efficiency Studies}: Detailed analysis of training time, memory usage, and computational efficiency at scale

    \item \textbf{Hybrid Optimization Strategies}: Investigation of combining multiple optimizers during different training phases

    \item \textbf{Auto-ML Integration}: Development of automated optimizer selection based on architecture and dataset characteristics
\end{itemize}

\subsection{Practical Applications}

Future work should explore:
\begin{itemize}
    \item \textbf{Production Deployment}: Large-scale evaluation in real-world search systems
    \item \textbf{Specialized Domains}: Medical, legal, and scientific document reranking
    \item \textbf{Multi-Modal Extensions}: Optimization strategies for vision-language retrieval tasks
\end{itemize}

\section{Practical Recommendations}

For practitioners working with cross-encoder reranking:

\begin{itemize}
    \item \textbf{Start with AdamW}: Use AdamW as a robust baseline for initial experiments
    \item \textbf{Consider Lion for Modern Architectures}: Explore Lion optimizer with appropriate learning rate tuning for recent transformer variants
    \item \textbf{Monitor Training Dynamics}: Track multiple metrics and loss curves for optimal checkpoint selection
    \item \textbf{Architecture-Specific Tuning}: Adjust optimizer parameters based on model architecture characteristics
\end{itemize}

\section{Conclusion}

This thesis demonstrates that optimizer choice significantly impacts cross-encoder reranking performance, with the effectiveness varying by architecture and configuration. The Lion optimizer shows particular promise for modern transformer architectures when properly tuned, while AdamW remains a reliable choice across different settings. These findings provide both practical guidance for practitioners and a foundation for future research in neural information retrieval optimization.

@@

We gratefully acknowledge Modal Labs for providing the cloud computing infrastructure and GPU resources (NVIDIA A100-80GB) that enabled this comprehensive experimental study.
